### 2018-05-23 conclusions.tex
changes before final submission

Identifying the total energy and response spectra metrics as decisive metrics in the comparisons between synthetics and observations is a key contribution to validation of ground motion simulations. This contributes to clarifying a standing question in the area of validation, and it provides an indication to simulators about where \change{to focus} in the search for improvements in their models.

>>>>

Identifying the total energy and response spectra metrics as decisive metrics in the comparisons between synthetics and observations is a key contribution to validation of ground motion simulations. This contributes to clarifying a standing question in the area of validation, and it provides an indication to simulators about where to focus in the search for improvements in their models.

### 2018-05-23 data-resources.tex
changes before final submission

Calculations, data processing, and some initial figures were done using R, the language and environment for statistical computing and graphics (\url{https://www.r-project.org}, last accessed April 2018), and the C5.0 package developed by \citet{Kuhn_2017_Manual}. Additional calculations and figures were done using MATLAB (\url{http://www.mathworks.com}, last accessed April 2018). Map figures were prepared using the Generic Mapping Tools, GMT (\url{http://gmt.soest.hawaii.edu/}, last accessed April 2018). Additional editing of figures was done using Adobe Illustrator (\url{http://www.adobe.com/Illustrator‎}, last accessed April 2018). The validation dataset used here was readily available to the authors and can be provided without restrictions upon request.

>>>>

Calculations, data processing, and some initial figures were done using R, the language and environment for statistical computing and graphics (\url{https://www.r-project.org}, last accessed May 2018), and the C5.0 package (\url{https://CRAN.R-project.org/package=C50}, last accessed May 2018). Additional calculations and figures were done using MATLAB (\url{http://www.mathworks.com}, last accessed April 2018). Map figures were prepared using the Generic Mapping Tools, GMT (\url{http://gmt.soest.hawaii.edu/}, last accessed May 2018). Additional editing of figures was done using Adobe Illustrator (\url{http://www.adobe.com/Illustrator‎}, last accessed May 2018). The validation dataset used here was readily available to the authors and can be provided without restrictions upon request.

### 2018-05-23 dataset.tex
changes before final submission

We select the simulation results and validation analysis done by \citet{Taborda_2014_BSSA} as our \textit{validation dataset}. In that study, the authors carried out deterministic simulations for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake using a finite-element approach. The simulations, done for a kinematic finite-fault model of the earthquake, were computed for a maximum frequency \fmaxeq{4} and a minimum shear wave velocity \vsmineq{200}. In total, \citet{Taborda_2014_BSSA} did three simulations, each for a different velocity model \citep[CVM-S4, CVM-H, CVM-H+GTL, see][]{Small_2017_SRL}. The modeling domain covered an area of \adomain{180}{135}{km} that included all the major sedimentary basins and other relevant geologic structures in the greater Los Angeles region. Their validation analysis consisted of comparisons with data recorded during the event at 336 ground motion monitoring stations, for the three components of motion (EW, NS, and UD). The simulation domain and the stations used in that study are shown in Figure \ref{fig:chino-hills}, and a sample of the validation results obtained by the authors using \citeauthor{Anderson_2004_Proc}'s approach is shown in Figure \ref{fig:ref-gof-maps}. This second figure, in particular, shows the spatial distribution of GOF scores (interpolated from the values at each station) for the comparison of the broadband synthetics and data, \change{for the three velocity models used by \citet{Taborda_2014_BSSA}. In each case, the results are the average of the GOF results for the three components of motion and the scores for the eleven metrics from Table \ref{tab:metrics}. A subsequent study by \citet{Taborda_2016_GJI} using multiple events found that the model CVM-S \citep[version 4.26.M01, see][]{Small_2017_SRL} to be the model that most consistently led to better simulation results. Here, however, we use the GOF scores obtained by \citet{Taborda_2014_BSSA} independently of the velocity models and/or the components of motion}.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-02}
    \caption{Validation results obtained by \citet{Taborda_2014_BSSA} in the form of GOF values across the region of interest obtained from comparisons between simulations for three different velocity models and data recorded at ground motion monitoring stations. \change{The distribution of the results at the bottom show the count of stations on the GOF scale defined by \citet{Anderson_2004_Proc} and the ranges of the validation categories: poor, fair, good and excellent.} The color version of this figure is available only in the electronic edition.}
    \label{fig:ref-gof-maps}
\end{figure*}

We refer to the set of 11 scores associated with a pair of signals \change{from the work done by \citet{Taborda_2014_BSSA}} as one of the \textit{data samples} that compose our dataset. Although at times we will make distinctions between the velocity models and the components of motion for visualization purposes, the clustering analysis to be described in the following section was done using all the data samples in the validation dataset. The motivation behind this choice was that the dataset, as a collection of GOF values, was independent of the simulations and serves here as a generic set of data samples for the purpose of identifying the correlations that exist between the different metrics in Table \ref{tab:metrics}. As such, given the simulations for each velocity model (3), the components of motion (3), and the number of stations used in the validation (336), gave us a large enough dataset, with 3,024 data samples. Figure \ref{fig:data-box-plot} illustrates the idea of a lack of dependence on the simulations by comparing the statistical distribution of the GOF scores of the simulations organized by velocity models and components. It is clear that although there are some differences between them, these are negligible. In other words, the statistical distribution of the data for each metric is about the same independently of model or component. Consequently, \change{and in order to use a dataset as large as possible, we thought it acceptable} to combine all the data samples into a single dataset.

>>>>

We select the simulation results and validation analysis done by \citet{Taborda_2014_BSSA} as our \textit{validation dataset}. In that study, the authors carried out deterministic simulations for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake using a finite-element approach. The simulations, done for a kinematic finite-fault model of the earthquake, were computed for a maximum frequency \fmaxeq{4} and a minimum shear wave velocity \vsmineq{200}. In total, \citet{Taborda_2014_BSSA} did three simulations, each for a different velocity model \citep[CVM-S4, CVM-H, CVM-H+GTL, see][]{Small_2017_SRL}. The modeling domain covered an area of \adomain{180}{135}{km} that included all the major sedimentary basins and other relevant geologic structures in the greater Los Angeles region. Their validation analysis consisted of comparisons with data recorded during the event at 336 ground motion monitoring stations, for the three components of motion (EW, NS, and UD). The simulation domain and the stations used in that study are shown in Figure \ref{fig:chino-hills}, and a sample of the validation results obtained by the authors using \citeauthor{Anderson_2004_Proc}'s approach is shown in Figure \ref{fig:ref-gof-maps}. This second figure, in particular, shows the spatial distribution of GOF scores (interpolated from the values at each station) for the comparison of the broadband synthetics and data, for the three velocity models used by \citet{Taborda_2014_BSSA}. In each case, the results are the average of the GOF results for the three components of motion and the scores for the eleven metrics from Table \ref{tab:metrics}. A subsequent study by \citet{Taborda_2016_GJI} using multiple events found that the model CVM-S \citep[version 4.26.M01, see][]{Small_2017_SRL} to be the model that most consistently led to better simulation results. Here, however, we use the GOF scores obtained by \citet{Taborda_2014_BSSA} independently of the velocity models and/or the components of motion.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-02}
    \caption{Validation results obtained by \citet{Taborda_2014_BSSA} in the form of GOF values across the region of interest obtained from comparisons between simulations for three different velocity models and data recorded at ground motion monitoring stations. The distribution of the results at the bottom show the count of stations on the GOF scale defined by \citet{Anderson_2004_Proc} and the ranges of the validation categories: poor, fair, good and excellent. The color version of this figure is available only in the electronic edition.}
    \label{fig:ref-gof-maps}
\end{figure*}

We refer to the set of 11 scores associated with a pair of signals from the work done by \citet{Taborda_2014_BSSA} as one of the \textit{data samples} that compose our dataset. Although at times we will make distinctions between the velocity models and the components of motion for visualization purposes, the clustering analysis to be described in the following section was done using all the data samples in the validation dataset. The motivation behind this choice was that the dataset, as a collection of GOF values, was independent of the simulations and serves here as a generic set of data samples for the purpose of identifying the correlations that exist between the different metrics in Table \ref{tab:metrics}. As such, given the simulations for each velocity model (3), the components of motion (3), and the number of stations used in the validation (336), gave us a large enough dataset, with 3,024 data samples. Figure \ref{fig:data-box-plot} illustrates the idea of a lack of dependence on the simulations by comparing the statistical distribution of the GOF scores of the simulations organized by velocity models and components. It is clear that although there are some differences between them, these are negligible. In other words, the statistical distribution of the data for each metric is about the same independently of model or component. Consequently, and in order to use a dataset as large as possible, we thought it acceptable to combine all the data samples into a single dataset.

### 2018-05-23 results.tex
changes before final submission

Some aspects of Fig.~\ref{fig:clusters} are worth discussing. \change{First of all, one must keep in mind that these are not typical correlation plots. How well defined the clusters are from each other or with respect to the metrics are important aspects to consider.} Although in most plots all four clusters are clearly distinguishable, there are others in which that is not the case. In the combinations in the 3-feature analysis, for instance, the excellent cluster has a limited presence. This is due to the influence of the lower C9 values (see Fig.~\ref{fig:data-box-plot}) on the correlations with C5 and C7. This is not necessarily a bad thing, as it helps to understand the different relationships between the various GOF metrics. In the best of cases, the combinations C1-C2 and C5-C8 in the 2- and 4-feature analysis, for instance, exhibit an almost perfect proportionality between the metrics. That means either one of the metrics in those combinations is redundant. Relationships with redundant features are those where knowledge about one of the features provides a direct view into the other. On the other hand, the combination C1-C7 in the 2-feature analysis presents an example of an irrelevant feature. We say C1 is irrelevant because it provides no insight about the outcome of C7 or that of the clusters. Identifying subspaces with redundant and irrelevant features is important because, on the one hand, the former help reduce the number of necessary features, while on the other hand, the latter can essentially be discarded because of their weak contribution to the decision making process \citep{Dy_2004_MLR}.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08}
	\caption{Accuracy of tree predictions in terms of the factor ($F_1$) from equation \ref{eq:f} as indicated by colored dots distributed with respect to the number of attributes (GOF metrics) as a function of: \textbf{(a)} the number of nodes, and \textbf{(b)} the depth of the trees. The rings around some of the dots indicate select trees used as reference in the text, other figures and tables. The color version of this figure is available only in the electronic edition.}
	\label{fig:nodes-depth}
\end{figure*}

@@

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of their topology on the right-hand side. T3 is the most complex of them. More complex trees tend to be deeper, have more nodes, and employ more attributes, all of which depends---in part---on the pruning process. More complex trees have higher $F_1$ values, but they may not necessarily be more practical. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of \change{the total energy (C4) and response spectra (C8) as key metrics. T2 adds the peak acceleration (C5), and T3 adds the peak velocity (C6)} to the previous metrics.

@@

The results in Table 6 highlight the fact that, for all trees, the metrics \change{for total energy (C4) and response spectra (C8) are consistently the most relevant ones in the decision algorithms. They are followed by the peak acceleration (C5) and peak velocity (C6). On the other end, the strong phase duration (C11) plays no role whatsoever in any of the trees, and the Arias integral (C1), energy integral (C2), and Arias intensity (C3) have only small to marginal participations. The remaining metrics---peak displacement (C7), Fourier spectrum (C9), and cross correlation (C10)}---have a somewhat significant participation, but only in the more complex trees.

An additional aspect worth highlighting here is the fact that the clustering and decision tree processes address the problem of correlation between the metrics in a natural way. For instance, one would expect \change{the Arias intensity (C3) and peak acceleration (C5), or the Arias intensity (C3) and total energy (C4)} to be related. Considering them all without distinction could lead to double weighing their influence, and assuming a 1-to-1 relationship may overstate their level of correlation. Instead, the process used here determines which of them has a better chance to predict the outcome in the presence of the other, and preserves that metric as part of the decision tree.

One can put this to test by removing one or more metrics and see how the participation of the metrics rearranges. Table \ref{tab:participation-extra} shows these participations for trees with similar topologies (i) when removing \change{the response spectrum (C8), and (ii) when removing both the peak acceleration (C5) and the response spectrum (C8). In the first case, the peak acceleration (C5) takes the place of the most determinant metric, followed by the total energy (C4) and peak velocity (C6), depending on the tree. In the second case, the peak velocity (C6) becomes the most dominant, followed by the Arias intensity (C3) and the total energy (C4), which vary in their level of participation depending on the tree. These additional alternatives emphasize the role of the total energy (C4) and the peak velocity (C6), and show that other metrics such as the peak acceleration (C5), peak displacement (C7) and Fourier spectrum (C9) are also relevant. They also reaffirm the low or null levels of contribution of the metrics associated with the shape of the Arias integral (C1), the energy integral (C2), and the cross correlation (C10) and strong phase duration (C11)}. All this is consistent with what we observed in Fig.~\ref{fig:boxed-clusters}.

That being said, a low participation does not necessarily mean that a given metric is not relevant at all. Arguably, the strong phase duration is highly regarded as an important parameter for strong motion records in engineering. What the results we present here mean is that, in the context of this particular set of 11 metrics, in order to classify whether a simulation result is poor, fair, good, or excellent, other metrics such as \change{the total energy (C4) are much more influential in the final result than the strong phase duration (C11)}. This, of course, is done under the assumption that one seeks to narrow the selection of metrics without loss of insight about the outcome of the validation process when compared to the outcome obtained with the whole suite of the eleven metrics used here.

In the end, the final selection of a preferred tree comes down to reducing complexity in the analysis without compromising the interpretation of results. We favor tree T1 because it is based only on three decision steps, and two GOF metrics: \change{the total energy (C4) and response spectra (C8)}.

>>>>

Some aspects of Fig.~\ref{fig:clusters} are worth discussing. First of all, one must keep in mind that these are not typical correlation plots. How well defined the clusters are from each other or with respect to the metrics are important aspects to consider. Although in most plots all four clusters are clearly distinguishable, there are others in which that is not the case. In the combinations in the 3-feature analysis, for instance, the excellent cluster has a limited presence. This is due to the influence of the lower C9 values (see Fig.~\ref{fig:data-box-plot}) on the correlations with C5 and C7. This is not necessarily a bad thing, as it helps to understand the different relationships between the various GOF metrics. In the best of cases, the combinations C1-C2 and C5-C8 in the 2- and 4-feature analysis, for instance, exhibit an almost perfect proportionality between the metrics. That means either one of the metrics in those combinations is redundant. Relationships with redundant features are those where knowledge about one of the features provides a direct view into the other. On the other hand, the combination C1-C7 in the 2-feature analysis presents an example of an irrelevant feature. We say C1 is irrelevant because it provides no insight about the outcome of C7 or that of the clusters. Identifying subspaces with redundant and irrelevant features is important because, on the one hand, the former help reduce the number of necessary features, while on the other hand, the latter can essentially be discarded because of their weak contribution to the decision making process \citep{Dy_2004_MLR}.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08}
	\caption{Accuracy of tree predictions in terms of the factor ($F_1$) from equation \ref{eq:f} as indicated by dots distributed with respect to the number of attributes (GOF metrics) as a function of: \textbf{(a)} the number of nodes, and \textbf{(b)} the depth of the trees. The rings around some of the dots indicate select trees used as reference in the text, other figures and tables. The color version of this figure is available only in the electronic edition.}
	\label{fig:nodes-depth}
\end{figure*}

@@

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of their topology on the right-hand side. T3 is the most complex of them. More complex trees tend to be deeper, have more nodes, and employ more attributes, all of which depends---in part---on the pruning process. More complex trees have higher $F_1$ values, but they may not necessarily be more practical. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of the total energy (C4) and response spectra (C8) as key metrics. T2 adds the peak acceleration (C5), and T3 adds the peak velocity (C6) to the previous metrics.

@@

The results in Table \ref{tab:participation} highlight the fact that, for all trees, the metrics for total energy (C4) and response spectra (C8) are consistently the most relevant ones in the decision algorithms. They are followed by the peak acceleration (C5) and peak velocity (C6). On the other end, the strong phase duration (C11) plays no role whatsoever in any of the trees, and the Arias integral (C1), energy integral (C2), and Arias intensity (C3) have only small to marginal participations. The remaining metrics---peak displacement (C7), Fourier spectrum (C9), and cross correlation (C10)---have a somewhat significant participation, but only in the more complex trees.

An additional aspect worth highlighting here is the fact that the clustering and decision tree processes address the problem of correlation between the metrics in a natural way. For instance, one would expect the Arias intensity (C3) and peak acceleration (C5), or the Arias intensity (C3) and total energy (C4) to be related. Considering them all without distinction could lead to double weighing their influence, and assuming a 1-to-1 relationship may overstate their level of correlation. Instead, the process used here determines which of them has a better chance to predict the outcome in the presence of the other, and preserves that metric as part of the decision tree.

One can put this to test by removing one or more metrics and see how the participation of the metrics rearranges. Table \ref{tab:participation-extra} shows these participations for trees with similar topologies (i) when removing the response spectrum (C8), and (ii) when removing both the peak acceleration (C5) and the response spectrum (C8). In the first case, the peak acceleration (C5) takes the place of the most determinant metric, followed by the total energy (C4) and peak velocity (C6), depending on the tree. In the second case, the peak velocity (C6) becomes the most dominant, followed by the Arias intensity (C3) and the total energy (C4), which vary in their level of participation depending on the tree. These additional alternatives emphasize the role of the total energy (C4) and the peak velocity (C6), and show that other metrics such as the peak acceleration (C5), peak displacement (C7) and Fourier spectrum (C9) are also relevant. They also reaffirm the low or null levels of contribution of the metrics associated with the shape of the Arias integral (C1), the energy integral (C2), and the cross correlation (C10) and strong phase duration (C11). All this is consistent with what we observed in Fig.~\ref{fig:boxed-clusters}.

That being said, a low participation does not necessarily mean that a given metric is not relevant at all. Arguably, the strong phase duration is highly regarded as an important parameter for strong motion records in engineering. What the results we present here mean is that, in the context of this particular set of 11 metrics, in order to classify whether a simulation result is poor, fair, good, or excellent, other metrics such as the total energy (C4) are much more influential in the final result than the strong phase duration (C11). This, of course, is done under the assumption that one seeks to narrow the selection of metrics without loss of insight about the outcome of the validation process when compared to the outcome obtained with the whole suite of the eleven metrics used here.

In the end, the final selection of a preferred tree comes down to reducing complexity in the analysis without compromising the interpretation of results. We favor tree T1 because it is based only on three decision steps, and two GOF metrics: the total energy (C4) and response spectra (C8).

### 2018-05-23 testing.tex
changes before final submission

We now test tree T1 on the original simulation results from \citet{Taborda_2014_BSSA}, for different velocity models and components of motion. In this case we no longer aggregate all data from the simulation but now look at individual simulations done with different velocity models, and the three components of motion separately, as it would normally be done during any ground motion simulation validation. The results obtained with the T1 validation algorithm are shown in Fig.~\ref{fig:res-gof-maps}. We should note that they are not supposed to be the same as those obtained by \citet{Taborda_2014_BSSA} because T1 no longer gives the same weight to all metrics but relies only on the two metrics C4 and C8. \change{A drawback of the T1 algorithm} is that because its outcome are GOF validation classes (poor, fair, good, excellent) as opposed to GOF validation scores (with values from 0 to 10), once one obtains the results for the validation process for different components of motion (as shown in the rows of Fig.~\ref{fig:res-gof-maps}), there is no natural way of taking averages as one would do for scores in a numerical scale \citep[i.e., as typically done when using][]{Anderson_2004_Proc}. Therefore, in order to combine results from the three components into a single validation classification for each station, we define the following rules:

@@

In our view, the T1 results are simpler to assimilate and equally informative. Looking at both plots in Fig.~\ref{fig:avg-gof-maps} it is possible to argue they lead to similar conclusions in respect to the overall validation of the simulation, and in respect to some particular areas and specific locations. Nonetheless, it must be said that we do not expect to see a 1-to-1 relationship\change{, as evidenced by the differences in the histograms when compared in equal terms (see dashed-line counts on the right-hand side corresponding to the count of the original results from the left-hand side histogram drawn next to the counts of the results from the T1 algorithm). Finding the best possible match} is the scope of a future work, whereas here our focus was on identifying the relevance of the different metrics. This is important because the original approach proposed by \citet{Anderson_2004_Proc} favors a uniform weighing of metrics, and the results shown here indicate this may not be the preferred strategy.

\begin{figure*}[th!]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-11}
	\caption{Comparison of GOF validation scores obtained using an 11-metric \citet{Anderson_2004_Proc}-type GOF scoring (left) and the T1 combination GOF validation classification (right). The top plots show maps with a distribution of the scores or classes outcome at each station. The contours are drawn for illustration purposes only. The bottom plots show histograms with the count of stations for each GOF score interval or validation class. \change{On the right-hand side, the results of both methods are compared by superimposing the results of the counts from using \citeauthor{Anderson_2004_Proc}'s method (dashed-line empty bars) next to those obtained using the T1 algorithm (filled bars).} The color version of this figure is available only in the electronic edition.}
	\label{fig:avg-gof-maps}
\end{figure*}

>>>>

We now test tree T1 on the original simulation results from \citet{Taborda_2014_BSSA}, for different velocity models and components of motion. In this case we no longer aggregate all data from the simulation but now look at individual simulations done with different velocity models, and the three components of motion separately, as it would normally be done during any ground motion simulation validation. The results obtained with the T1 validation algorithm are shown in Fig.~\ref{fig:res-gof-maps}. We should note that they are not supposed to be the same as those obtained by \citet{Taborda_2014_BSSA} because T1 no longer gives the same weight to all metrics but relies only on the two metrics C4 and C8. A drawback of the T1 algorithm is that because its outcome are GOF validation classes (poor, fair, good, excellent) as opposed to GOF validation scores (with values from 0 to 10), once one obtains the results for the validation process for different components of motion (as shown in the rows of Fig.~\ref{fig:res-gof-maps}), there is no natural way of taking averages as one would do for scores in a numerical scale \citep[i.e., as typically done when using][]{Anderson_2004_Proc}. Therefore, in order to combine results from the three components into a single validation classification for each station, we define the following rules:

@@

In our view, the T1 results are simpler to assimilate and equally informative. Looking at both plots in Fig.~\ref{fig:avg-gof-maps} it is possible to argue they lead to similar conclusions in respect to the overall validation of the simulation, and in respect to some particular areas and specific locations. Nonetheless, it must be said that we do not expect to see a 1-to-1 relationship, as evidenced by the differences in the histograms when compared in equal terms (see dashed-line counts on the right-hand side corresponding to the count of the original results from the left-hand side histogram drawn next to the counts of the results from the T1 algorithm). Finding the best possible match is the scope of a future work, whereas here our focus was on identifying the relevance of the different metrics. This is important because the original approach proposed by \citet{Anderson_2004_Proc} favors a uniform weighing of metrics, and the results shown here indicate this may not be the preferred strategy.

\begin{figure*}[th!]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-11}
	\caption{Comparison of GOF validation scores obtained using an 11-metric \citet{Anderson_2004_Proc}-type GOF scoring (left) and the T1 combination GOF validation classification (right). The top plots show maps with a distribution of the scores or classes outcome at each station. The contours are drawn for illustration purposes only. The bottom plots show histograms with the count of stations for each GOF score interval or validation class. On the right-hand side, the results of both methods are compared by superimposing the results of the counts from using \citeauthor{Anderson_2004_Proc}'s method (dashed-line empty bars) next to those obtained using the T1 algorithm (filled bars). The color version of this figure is available only in the electronic edition.}
	\label{fig:avg-gof-maps}
\end{figure*}
